Online {\IterComp}\\Guided by Low Overhead Profiling

### 2017-05-09 cgo18mscr.tex
minor change: organising line breaks

\begin{document}
%\title{A work-based metric for efficiently applying online iterative optimisation}
\title{Online {\IterComp}\\Guided by Low Overhead Profiling}
%\titlenote{Produces the permission block, and
%  copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as
%  \texttt{acmart.pdf} document}

@@

A well known compilation technique that
addresses this challenge is {\itercomp}.
{\Itercomp} has the ability to adapt to new
platforms, program and workload while still
having a systematic and simple optimisation process.
{\Itercomp} repeatedly evaluates a large number
of compiler optimisations until the best combination is
found for a particular program~\cite{fursin07,chen10}.
The main challenge concerning {\itercomp} is the need for quickly
exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.

@@

Most of real world applications are complex enough 
so that a single input case does not capture the whole
range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
Because programs can exhibit behaviours that differ greatly depending on the input,
%This is something that is well known when writing test
%cases for correctness.
%We can extend the same argument when optimising the code for performance.
using a single input for {\itercomp} can produce poor performance when executed with different inputs.

@@

%The hypothesis is that, given two optimisations $o_1$ and $o_2$,
%a program compiled with optimisation $o_2$ would \textit{consistently} perform better
%than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$,
%if it performs more \textit{work} per unit of time when compiled with $o_2$
%instead of $o_1$. 
By measuring the amount of \textit{work} done per unit of time we reduce
the impact of input-dependent aspects and focus instead on the efficiency
of the optimised program.
For this metric, the main challenge is to precisely define what
represents \textit{work}.

%For that purpose, we intend to use profiling
%techniques to measure the input size and amount of computation
%performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.

%In addition to being useful for speeding up {\itercomp} across a large
%number of different inputs, this metric based on comparing work-based efficiencies could
%potentially be used for applying {\itercomp} \textit{online}. 
%However, the overhead for estimating the amount of \textit{work} must be low
%enough for the performance gains to be beneficial.

@@

%By construction, 
%the \textit{symbolic experssions} of a basic block, i.e., 
%the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block,
%has the following properties:
%\begin{enumerate}
%\item Probes in the increment set either (post-)dominates or is (post-)dominated by the given basic block;
%\item Probes in the decrement set are (post-)dominated by the probes in the increment set;
%\item Probes in the decrement set and the given basic block are in exclusive paths
%      from the virtual node of the CFG to the (post-)dominators in the increment set;
%\end{enumerate}
%From property (1) it follows that whenever the given basic block is executed,
%exactly one of the probes in the increment set is executed.
%From properties (2) and (3), it follows that whenever a probe in the decrement set
%is executed, although the given basic block is not executed, one of the probes
%in the decrement set is executed.

@@

%Define an extended DAG of a CFG as a directed graph which may contain cycles
%if they have known constant trip counts.

@@

%
%LLVM has a Scalar Evolution Analysis which is used primarily to analyse expressions
%involving induction variables in loops~\cite{pop05}.
%An induction variable is a variable that is increased or decreased by a fixed amount
%on every iteration of a loop or is a linear function of another induction variable.
%
%This Scalar Evolution Analysis provides a way to compute \textit{small} constant
%trip counts (where \textit{small} is defined by a threshold).

@@

We model the relaxation as a 0-1 Knapsack problem:
\[
%\textrm{maximise } \sum_{i=0}^{k} f(I_i)c(I_i)x_i
\textrm{maximise } \sum_{i=0}^{k} f(P_i)x_i
\]
\[
\textrm{subject to } \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \textrm{ and } x_i\in\{0,1\}
\]
%$c(I_i)$ is the relative overhead of the instrumentation on the instrumented basic block,
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$
,$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative
to the minimum work value possible to compute in the extended DAG,
and $x_i$ denotes the probes selected for removal.
Because the percentage error is computed based on the path with the minimum
amount of work, $\varepsilon(P_i)$ represents the maximum error possible
that would be incurred when removing probe $P_i$.
Furthermore, by constraining the percentage error of every extended DAG below a
given threshold we guarantee that the final error of the relaxation will always
be bounded by the threshold.

@@

%{\Itercomp} is essentially based on repeatedly trying
%out a large number of compiler optimisations until the best
%combination of compiler optimisations is found for a particular program.
%Our main goal is to speed up {\itercomp} while targeting
%optimal performance across large input datasets.

@@

\begin{table}[h]
\centering
\scalebox{.8}{
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Program} & \textbf{LOC}    & \textbf{Input file size}            & \textbf{Input description}              \\ \hline % Domain
bitcount      & 460    &  -                         & Numbers: random                \\ \hline
qsort         & 154    & 32K-1.8M                   & 3D coordinates                 \\ \hline
%dijkstra      & 163    & 0.06K-4.3M                 & Adjacency matrices             \\ \hline
%patricia      & 290    & 0.6K-1.9M                  & IP and mask pairs              \\ \hline
jpeg\_d       & 13501  & 3.6K-1.5M                  & JPEG images                    \\ \hline
jpeg\_c       & 14014  & 16K-137M                   & PPM images                     \\ \hline
tiff2bw       & 15477  & \multirow{4}{*}{9K-137M}   & \multirow{4}{*}{TIFF images}   \\ \cline{1-2}
tiff2rgba     & 15424  &                            &                                \\ \cline{1-2}
tiffdither    & 15399  &                            &                                \\ \cline{1-2}
tiffmedian    & 15870  &                            &                                \\ \hline
susan\_c      & 1376   & \multirow{3}{*}{12K-46M}   & \multirow{3}{*}{PGM images}    \\ \cline{1-2}
susan\_e      & 1376   &                            &                                \\ \cline{1-2}
susan\_s      & 1376   &                            &                                \\ \hline
%mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
adpcm\_c      & 210    & 167K-36M                   & WAVE audios                    \\ \hline
adpcm\_d      & 211    & 21K-8.8M                   & ADPCM audios                   \\ \hline
%gsm           & 3806   & 83K-18M                    & Sun/NeXT audios                \\ \hline
%ghostscript   & 99869  & 11K-43M                    & Postscript files               \\ \hline
%ispell        & 6522   & \multirow{3}{*}{0.1K-42M}  & \multirow{3}{*}{Text files}    \\ \cline{1-2}
%rsynth        & 4111   &                            &                                \\ \cline{1-2}
rsynth        & 4111   &                            &  Text files                    \\ \hline %%%%%% copied line
%stringsearch  & 338    &                            &                                \\ \hline
%blowfish\_e   & 863    & 0.6K-35M                   & Files of any format            \\ \hline
%blowfish\_d   & 863    & 0.6K-35M                   & Encrypted files                \\ \hline
%pgp\_e        & 19575  & 0.6K-35M                   & Files of any format            \\ \hline
%pgp\_d        & 19575  & 0.4K-18M                   & Encrypted files                \\ \hline
%rijndael\_e   & 952    & 0.6K-35M                   & Files of any format            \\ \hline
%rijndael\_d   & 952    & 0.7K-35M                   & Encrypted files                \\ \hline
sha           & 197    & 0.6K-35M                   & Files of any format            \\ \hline
%CRC32         & 130    & 0.6K-35M                   & Files of any format            \\ \hline
%bzip2e        & 5125   & 0.7K-57M                   & Files of any format            \\ \hline
%bzip2d        & 5125   & 0.2K-25M                   & Compressed files               \\ \hline
\end{tabular}
}
\caption{Description of the KDataSets with 1000 inputs for each benchmark (Chen~\etal~\cite{chen10,chen12a}).}
\label{tab:kdatasets}
\end{table}

@@

%In this section we discuss the three baseline implementations that will
%be used for assessing the quality of the proposed metric.

@@

%Most of the existing {\itercomp} studies find the
%best optimisation though repeated runs on the same input.
%Although this approach will usually lead to sub-optimal performance
%across large input datasets, it provides a good baseline when considering
%the complexity and resonably low compile-time regarding
%iterative optimising compilers.
%If $M$ is the total number of combinations of compiler optimisations,
%this approach requires $O(M)$ runs of the program being optimised.
%We can consider two main scenarios:
%\textit{($i$ - best case scenario)}
%after selecting the optimisation over each individual input,
%consider the one with best performance across the whole input dataset;
%\textit{($ii$ - expected scenario)}
%after selecting the optimisation over each individual input,
%consider the average case of their performance across the whole input dataset.

@@

%Recent work on {\itercomp} have been targeting optimisation
%across multiple inputs~\cite{fursin07,chen10,chen12a}.
%If $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations,
%they perform a total of $O(NM)$ runs of the program being optimised.
%Similarly to what have been discussed previously,
%there are different ways for how to determine the optimal
%combination of compiler optimisations across multiple inputs.
%It is possible to tune the selection of the program-optimal combination
%to minimise risk or to maximise average performance.
%A compromise-based selection criterion could be to maximise average speedup
%with minimised variance.

@@

%While the previous two baselines addresses two opposite aspects of {\itercomp},
%namely, compile-time efficiency and performance of the generated code,
%we also intend to compare against IPC as the competing baseline.
%The main reason for comparing against IPC is because it has been proposed as a metric
%for comparing the performance of two optimisations running on two distinct inputs.
%If $M$ is the total number of combinations of compiler optimisations,
%this approach requires $O(M)$ runs of the program being optimised.

@@

%\subsubsection{Contribution of individual optimisation passes}
%
%Figure~\ref{fig:flagsfreq} shows an aggregated view of the final combination
%of compiler optimisations that were selected by {\itercomp} search.
%The figure presents the individual optimisation passes with at least 1\% of
%frequency in the selected combination of compiler optimisations that improved
%the performance over {\texttt{-O3}}.
%
%\begin{figure*}[htb]
%    \centering
%    \includegraphics[width=\textwidth]{figs/flagsfreq.pdf}
%    \caption{Frequency of individual optimisation passes on the final selected 
%             compiler optimisations of the {\itercomp} search over
%             all benchmarks.}
%    \label{fig:flagsfreq}
%\end{figure*}

>>>>

\begin{document}
\title{Online {\IterComp}\\Guided by Low Overhead Profiling}
%\titlenote{Produces the permission block, and
%  copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as
%  \texttt{acmart.pdf} document}

@@

A well known compilation technique that addresses this challenge is {\itercomp}.
{\Itercomp} has the ability to adapt to new platforms, program and workload while still having a systematic and simple optimisation process.
It works by repeatedly evaluating a large number of compiler optimisations until the best combination is found for a particular program~\cite{fursin07,chen10}.
The main challenge concerning {\itercomp} is the need for efficiently exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.

@@

Most of real world applications are complex enough so that a single input case does not capture the whole range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
Because programs can exhibit behaviours that differ greatly depending on the input,
%This is something that is well known when writing test cases for correctness.
%We can extend the same argument when optimising the code for performance.
using a single input for {\itercomp} can produce poor performance when executed with different inputs.

@@

%The hypothesis is that, given two optimisations $o_1$ and $o_2$, a program compiled with optimisation $o_2$ would \textit{consistently} perform better than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$, if it performs more \textit{work} per unit of time when compiled with $o_2$ instead of $o_1$. 
By measuring the amount of \textit{work} done per unit of time we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
For this metric, the main challenge is to precisely define what represents \textit{work}.

%For that purpose, we intend to use profiling techniques to measure the input size and amount of computation performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.

%In addition to being useful for speeding up {\itercomp} across a large number of different inputs, this metric based on comparing work-based efficiencies could
%potentially be used for applying {\itercomp} \textit{online}.
%However, the overhead for estimating the amount of \textit{work} must be low enough for the performance gains to be beneficial.

@@

%By construction, the \textit{symbolic experssions} of a basic block, i.e., the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block, has the following properties:
%\begin{enumerate}
%\item Probes in the increment set either (post-)dominates or is (post-)dominated by the given basic block;
%\item Probes in the decrement set are (post-)dominated by the probes in the increment set;
%\item Probes in the decrement set and the given basic block are in exclusive paths from the virtual node of the CFG to the (post-)dominators in the increment set;
%\end{enumerate}
%From property (1) it follows that whenever the given basic block is executed, exactly one of the probes in the increment set is executed.
%From properties (2) and (3), it follows that whenever a probe in the decrement set is executed, although the given basic block is not executed, one of the probes in the decrement set is executed.

@@

%Define an extended DAG of a CFG as a directed graph which may contain cycles if they have known constant trip counts.

@@

%
%LLVM has a Scalar Evolution Analysis which is used primarily to analyse expressions involving induction variables in loops~\cite{pop05}.
%An induction variable is a variable that is increased or decreased by a fixed amount on every iteration of a loop or is a linear function of another induction variable.
%
%This Scalar Evolution Analysis provides a way to compute \textit{small} constant trip counts (where \textit{small} is defined by a threshold).

@@

We model the relaxation as a 0-1 Knapsack problem:
\[
%\textrm{maximise } \sum_{i=0}^{k} f(I_i)c(I_i)x_i
\textrm{maximise } \sum_{i=0}^{k} f(P_i)x_i
\]
\[
\textrm{subject to } \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \textrm{ and } x_i\in\{0,1\}
\]
%$c(I_i)$ is the relative overhead of the instrumentation on the instrumented basic block,
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the extended DAG, and $x_i$ denotes the probes selected for removal.
Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
Furthermore, by constraining the percentage error of every extended DAG below a given threshold we guarantee that the final error of the relaxation will always be bounded by the threshold.

@@

%{\Itercomp} is essentially based on repeatedly trying out a large number of compiler optimisations until the best combination of compiler optimisations is found for a particular program.
%Our main goal is to speed up {\itercomp} while targeting optimal performance across large input datasets.

@@

\begin{table}[h]
\centering
\scalebox{.8}{
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Program} & \textbf{LOC}    & \textbf{Input file size}            & \textbf{Input description}              \\ \hline % Domain
%bitcount      & 460    &  -                         & Numbers: random                \\ \hline
qsort         & 154    & 32K-1.8M                   & 3D coordinates                 \\ \hline
%dijkstra      & 163    & 0.06K-4.3M                 & Adjacency matrices             \\ \hline
%patricia      & 290    & 0.6K-1.9M                  & IP and mask pairs              \\ \hline
jpeg\_d       & 13501  & 3.6K-1.5M                  & JPEG images                    \\ \hline
jpeg\_c       & 14014  & 16K-137M                   & PPM images                     \\ \hline
tiff2bw       & 15477  & \multirow{4}{*}{9K-137M}   & \multirow{4}{*}{TIFF images}   \\ \cline{1-2}
tiff2rgba     & 15424  &                            &                                \\ \cline{1-2}
tiffdither    & 15399  &                            &                                \\ \cline{1-2}
tiffmedian    & 15870  &                            &                                \\ \hline
susan\_c      & 1376   & \multirow{3}{*}{12K-46M}   & \multirow{3}{*}{PGM images}    \\ \cline{1-2}
susan\_e      & 1376   &                            &                                \\ \cline{1-2}
susan\_s      & 1376   &                            &                                \\ \hline
%mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
%lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
adpcm\_c      & 210    & 167K-36M                   & WAVE audios                    \\ \hline
adpcm\_d      & 211    & 21K-8.8M                   & ADPCM audios                   \\ \hline
%gsm           & 3806   & 83K-18M                    & Sun/NeXT audios                \\ \hline
%ghostscript   & 99869  & 11K-43M                    & Postscript files               \\ \hline
%ispell        & 6522   & \multirow{3}{*}{0.1K-42M}  & \multirow{3}{*}{Text files}    \\ \cline{1-2}
%rsynth        & 4111   &                            &                                \\ \cline{1-2}
rsynth        & 4111   &                            &  Text files                    \\ \hline %%%%%% copied line
%stringsearch  & 338    &                            &                                \\ \hline
%blowfish\_e   & 863    & 0.6K-35M                   & Files of any format            \\ \hline
%blowfish\_d   & 863    & 0.6K-35M                   & Encrypted files                \\ \hline
%pgp\_e        & 19575  & 0.6K-35M                   & Files of any format            \\ \hline
%pgp\_d        & 19575  & 0.4K-18M                   & Encrypted files                \\ \hline
%rijndael\_e   & 952    & 0.6K-35M                   & Files of any format            \\ \hline
%rijndael\_d   & 952    & 0.7K-35M                   & Encrypted files                \\ \hline
sha           & 197    & 0.6K-35M                   & Files of any format            \\ \hline
%CRC32         & 130    & 0.6K-35M                   & Files of any format            \\ \hline
%bzip2e        & 5125   & 0.7K-57M                   & Files of any format            \\ \hline
%bzip2d        & 5125   & 0.2K-25M                   & Compressed files               \\ \hline
\end{tabular}
}
\caption{Description of the KDataSets with 1000 inputs for each benchmark (Chen~\etal~\cite{chen10,chen12a}).}
\label{tab:kdatasets}
\end{table}

@@

%In this section we discuss the three baseline implementations that will be used for assessing the quality of the proposed metric.

@@

%Most of the existing {\itercomp} studies find the best optimisation though repeated runs on the same input.
%Although this approach will usually lead to sub-optimal performance across large input datasets, it provides a good baseline when considering the complexity and resonably low compile-time regarding iterative optimising compilers.
%If $M$ is the total number of combinations of compiler optimisations, this approach requires $O(M)$ runs of the program being optimised.
%We can consider two main scenarios:
%\textit{($i$ - best case scenario)} after selecting the optimisation over each individual input, consider the one with best performance across the whole input dataset;
%\textit{($ii$ - expected scenario)} after selecting the optimisation over each individual input, consider the average case of their performance across the whole input dataset.

@@

%Recent work on {\itercomp} have been targeting optimisation across multiple inputs~\cite{fursin07,chen10,chen12a}.
%If $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
%Similarly to what have been discussed previously, there are different ways for how to determine the optimal combination of compiler optimisations across multiple inputs.
%It is possible to tune the selection of the program-optimal combination to minimise risk or to maximise average performance.
%A compromise-based selection criterion could be to maximise average speedup with minimised variance.

@@

%While the previous two baselines addresses two opposite aspects of {\itercomp}, namely, compile-time efficiency and performance of the generated code, we also intend to compare against IPC as the competing baseline.
%The main reason for comparing against IPC is because it has been proposed as a metric for comparing the performance of two optimisations running on two distinct inputs.
%If $M$ is the total number of combinations of compiler optimisations, this approach requires $O(M)$ runs of the program being optimised.

@@

%\subsubsection{Contribution of individual optimisation passes}
%
%Figure~\ref{fig:flagsfreq} shows an aggregated view of the final combination of compiler optimisations that were selected by {\itercomp} search.
%The figure presents the individual optimisation passes with at least 1\% of frequency in the selected combination of compiler optimisations that improved the performance over {\texttt{-O3}}.
%
%\begin{figure*}[htb]
%    \centering
%    \includegraphics[width=\textwidth]{figs/flagsfreq.pdf}
%    \caption{Frequency of individual optimisation passes on the final selected 
%             compiler optimisations of the {\itercomp} search over
%             all benchmarks.}
%    \label{fig:flagsfreq}
%\end{figure*}
